---
title: "Missing-Data Imputation in NHANES 2017-2018: Controlled Benchmarking and Real-Data Application"
author: "Laurin Seelig, Luca Meier, Ravidu Nakandalage"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)
knitr::opts_knit$set(root.dir = here::here())
```

```{r packages-and-helpers, include=FALSE}
required_pkgs <- c("dplyr", "readr", "tidyr", "ggplot2", "knitr", "mice", "broom", "here")
missing_pkgs  <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop(
    "Missing required package(s): ",
    paste(missing_pkgs, collapse = ", "),
    ". Install these packages before rendering."
  )
}

library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(knitr)
library(mice)
library(broom)
library(here)

safe_read_csv <- function(path) {
  if (!file.exists(path)) { warning("File not found: ", path); return(NULL) }
  readr::read_csv(path, show_col_types = FALSE)
}
```

# Introduction

Missing data are ubiquitous in observational health studies and may strongly influence both predictive performance and inferential validity. In complex survey datasets such as NHANES, missingness can arise from item non-response, laboratory non-completion, or measurement constraints. If not handled rigorously, these mechanisms can produce biased estimates, reduced precision, and potentially misleading substantive conclusions.

This project investigates missing-data handling using two complementary perspectives:

1. **Controlled benchmarking** on complete-case reference data with artificially induced missingness under pre-specified mechanisms (MCAR and MAR) and missingness levels (10%, 20%, 30%), using **8 simulation repetitions** per setting (48 datasets in total).
2. **Real-data application** on NHANES 2017–2018 data, where the true missing values are unknown and conclusions therefore depend on assumptions.

The primary objective is to compare complete-case analysis, single imputation approaches (mean/mode, kNN, missForest), and multiple imputation (MICE) with respect to both **imputation accuracy** and **downstream regression inference** for glycated haemoglobin (HbA1c).

# Theory

## Missing-data mechanisms

Let $R$ denote a missingness indicator and $Y$ the data matrix partitioned into observed and missing components.

- **MCAR (Missing Completely At Random):**  
  $P(R \mid Y_{\text{obs}}, Y_{\text{mis}}) = P(R)$.  
  Missingness is independent of both observed and unobserved values. Complete-case analysis is unbiased under MCAR, but this assumption is rarely plausible in health data.

- **MAR (Missing At Random):**  
  $P(R \mid Y_{\text{obs}}, Y_{\text{mis}}) = P(R \mid Y_{\text{obs}})$.  
  Missingness may depend on observed covariates but not on the unobserved values after conditioning. Most modern imputation methods assume MAR.

- **MNAR (Missing Not At Random):**  
  Missingness depends on unobserved values even after conditioning on observed data. MNAR cannot be diagnosed from observed data alone and requires substantive assumptions or sensitivity analyses such as delta-adjustment or pattern-mixture models.

## Methods considered

- **Complete-case analysis (CCA):** discards all observations with any missing value. Valid under MCAR (or specific model structures), but reduces effective sample size and can introduce bias under MAR.
- **Mean/mode imputation:** replaces missing values with the variable mean (continuous) or mode (categorical). Simple and fast, but ignores variability in the missing data, attenuates correlations, and underestimates uncertainty in downstream analyses.
- **k-Nearest Neighbours (kNN):** imputes missing values from the $k$ most similar observed records. Captures local data structure but may struggle under higher missingness or sparse overlap ($k = 5$ used here).
- **missForest:** an ensemble-based non-parametric method using random forests to iteratively predict missing values. Handles non-linear relationships and mixed variable types (40 trees used here).
- **Multiple imputation by chained equations (MICE):** generates $m$ complete datasets by iteratively regressing each incomplete variable on all others, then pools estimates using Rubin's rules. Explicitly propagates imputation uncertainty into standard errors and confidence intervals ($m = 8$, 5 iterations used here).

## Evaluation principles

Two complementary performance criteria are used:

1. **Imputation accuracy** at artificially masked cells:
   - Normalised RMSE (NRMSE = RMSE / SD of observed variable) for continuous variables; values below 1.0 indicate the method outperforms naive mean imputation.
   - Proportion of falsely classified values (PFC) for categorical variables.

2. **Inferential performance** in the downstream model:
$$\text{HbA1c} \sim \text{age} + \text{sex} + \text{BMI} + \text{systolic\_bp} + \text{smoking\_status} + \text{income\_ratio}$$
   summarised by coefficient bias (estimate minus reference), empirical standard deviation, model-based standard errors, 95% confidence-interval coverage, and interval width.

# Data

The analysis is based on NHANES 2017–2018 and prepared using scripted extraction and cleaning from CDC XPT files. The final analysis dataset contains $n = 2{,}000$ adult participants (age $\geq$ 20 years) and seven variables relevant to metabolic health.

```{r load-prepared-data}
nhanes_analysis     <- safe_read_csv("data/nhanes_analysis.csv")
missing_var_summary <- safe_read_csv("outputs/missingness_variable_summary.csv")
missing_desc        <- safe_read_csv("outputs/missingness_descriptive_table.csv")
```

```{r data-overview-table}
if (!is.null(nhanes_analysis)) {
  data_overview <- tibble::tibble(
    Metric = c("Number of observations", "Number of variables"),
    Value  = c(nrow(nhanes_analysis), ncol(nhanes_analysis))
  )
  knitr::kable(data_overview, caption = "Table 1: Overview of the prepared NHANES analysis dataset.")
}
```

```{r missingness-summary-table}
if (!is.null(missing_var_summary)) {
  knitr::kable(
    missing_var_summary,
    col.names = c("Variable", "N missing", "N total", "Missing (%)"),
    caption   = "Table 2: Variable-level missingness in the prepared NHANES analysis dataset."
  )
}
```

The income-to-poverty ratio has the highest proportion of missing values (19.4%), followed by systolic blood pressure (10.5%), HbA1c (9.6%), and BMI (7.1%). Age, sex, and smoking status are fully observed.

```{r fig-missing-pct, echo=FALSE, fig.cap="Figure 1: Percentage of missing values by variable in the prepared NHANES analysis dataset."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/missingness_by_variable_pct.png")
)
```

```{r fig-missing-pattern, echo=FALSE, fig.cap="Figure 2: Pairwise missingness structure — percentage of observations in which both variables are simultaneously missing."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/missingness_pattern_plot.png")
)
```

The co-occurrence of missingness among BMI, HbA1c, systolic blood pressure, and income-to-poverty ratio (with up to 9.6% joint missingness for HbA1c and BMI) suggests a structured rather than purely random pattern. This is inconsistent with MCAR and motivates the use of model-based imputation methods under the MAR assumption.

# Methods

## Controlled benchmarking design

A complete-case reference subset ($n = 1{,}000$) was used to induce artificial missingness in predictors at 10%, 20%, and 30% under two mechanisms:

- **MCAR:** random masking, independent of all data.
- **MAR:** logistic masking probabilities dependent on observed age, sex, and BMI.

For each of the 6 mechanism–level combinations, **8 repetitions** were generated, yielding **48 simulated datasets** in total. Masked-cell indices and true values were retained for accuracy evaluation.

```{r simulation-manifest}
sim_manifest <- safe_read_csv("sim_data/simulation_manifest.csv")
if (!is.null(sim_manifest)) {
  sim_overview <- sim_manifest %>%
    group_by(mechanism, missing_rate) %>%
    summarise(
      n_repetitions     = n(),
      mean_masked_cells = mean(n_masked_cells),
      .groups = "drop"
    )
  knitr::kable(
    sim_overview,
    col.names = c("Mechanism", "Missing rate", "Repetitions", "Mean masked cells"),
    caption   = "Table 3: Simulation design summary by mechanism and missingness level."
  )
}
```

## Imputation methods

Five approaches were applied to each simulated dataset:

| Method | Type | Key parameters |
|---|---|---|
| Complete-case analysis | — | Listwise deletion |
| Mean/mode imputation | Single | Mean (continuous), mode (categorical) |
| kNN | Single | $k = 5$ |
| missForest | Single | 40 trees |
| MICE | Multiple | $m = 8$, 5 iterations, PMM / logreg / polyreg |

For MICE, variable-type-appropriate methods were used: predictive mean matching (`pmm`) for continuous variables, logistic regression (`logreg`) for binary factors, and polytomous regression (`polyreg`) for unordered factors. Final inference was obtained by pooling via Rubin's rules.

## Downstream inferential model

After imputation, a linear model for HbA1c was fit in each repetition. For MICE, results were pooled using `with()` and `pool()` from the `mice` package. The reference coefficients were estimated from the complete-case reference subset (before missingness was introduced).

## Real-data application

For the observed NHANES data, the same model was applied to (i) complete cases only and (ii) a MICE analysis (20 imputations, 10 iterations), yielding pooled coefficients and confidence intervals. This analysis is interpretive, because the true missing values are unknown.

# Results

## Controlled benchmarking: imputation accuracy

```{r load-evaluation-outputs}
acc_rep   <- safe_read_csv("outputs/imputation_accuracy_by_repetition.csv")
acc_sum   <- safe_read_csv("outputs/imputation_accuracy_summary.csv")
model_rep <- safe_read_csv("outputs/model_estimates_by_repetition.csv")
model_sum <- safe_read_csv("outputs/model_performance_summary.csv")
```

```{r fig-accuracy-numeric, echo=FALSE, fig.cap="Figure 3: Imputation accuracy for continuous variables (NRMSE) across methods, missingness mechanisms (MCAR / MAR), and missingness levels (10-30%). Lower values indicate better reconstruction; values below 1.0 outperform naive mean imputation."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/accuracy_numeric_methods.png")
)
```

**missForest consistently achieved the lowest NRMSE** across continuous variables and both mechanisms. For example, at MAR 30% for systolic blood pressure, missForest reached NRMSE = 1.00 compared to kNN (1.06), mean/mode (1.04), and MICE (1.33). For BMI under MCAR at all levels, missForest ranged from NRMSE = 0.96–0.97, while mean/mode remained close to 1.0 (i.e., no improvement over naive imputation) and MICE showed the highest error (NRMSE = 1.35–1.38). These patterns held under both MCAR and MAR, with errors slightly higher under MAR, consistent with the more difficult missingness structure.

MICE showed notably higher NRMSE than competing methods across most continuous variables. This is expected: MICE is designed to preserve the distribution and uncertainty of imputed values rather than minimise pointwise reconstruction error, which is measured by NRMSE.

```{r table-accuracy-summary}
if (!is.null(acc_sum)) {
  acc_main <- acc_sum %>%
    filter(metric == "nrmse", variable %in% c("bmi", "systolic_bp", "income_ratio")) %>%
    mutate(
      mean_value   = round(mean_value, 3),
      missing_rate = paste0(missing_rate * 100, "%")
    ) %>%
    select(Method = method, Mechanism = mechanism,
           `Missing rate` = missing_rate, Variable = variable,
           `Mean NRMSE` = mean_value)
  knitr::kable(
    acc_main,
    caption = "Table 4: NRMSE summary for selected continuous variables across methods, mechanisms, and missingness levels (complete-case excluded as no reconstruction is performed)."
  )
}
```

```{r fig-accuracy-categorical, echo=FALSE, fig.cap="Figure 4: Imputation accuracy for categorical variables (PFC = proportion falsely classified) across methods, missingness mechanisms, and levels. Lower values indicate fewer misclassifications."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/accuracy_categorical_methods.png")
)
```

For categorical variables (sex and smoking status), differences between methods were considerably smaller. PFC values ranged from approximately 0.38 to 0.53 across all methods and settings, with no consistent winner. missForest showed marginally lower PFC in several settings (e.g., smoking status MCAR 10%: PFC = 0.38 vs. mean/mode = 0.41), while mean/mode imputation occasionally performed worst due to its tendency to collapse predictions to the modal category. This reflects the difficulty of imputing binary variables that are close to 50/50 in prevalence.

## Controlled benchmarking: inferential performance

```{r fig-inference-bias, echo=FALSE, fig.cap="Figure 5: Mean coefficient bias (estimate minus reference) across methods, mechanisms, and missingness levels. Values closer to zero indicate better inferential performance."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/inference_bias_methods.png")
)
```

```{r fig-inference-coverage, echo=FALSE, fig.cap="Figure 6: Empirical 95% confidence-interval coverage across methods. The dashed line indicates the nominal 95% target. Coverage below this line indicates under-coverage (anti-conservative intervals)."}
knitr::include_graphics(
  file.path(knitr::opts_knit$get("root.dir"), "figures/inference_coverage_methods.png")
)
```

```{r table-inference-summary}
if (!is.null(model_sum)) {
  infer_main <- model_sum %>%
    filter(term != "(Intercept)") %>%
    mutate(
      mean_bias    = round(mean_bias, 4),
      ci_coverage  = round(ci_coverage, 3),
      avg_ci_width = round(avg_ci_width, 3),
      missing_rate = paste0(missing_rate * 100, "%")
    ) %>%
    arrange(term, mechanism, missing_rate, method) %>%
    select(Method = method, Mechanism = mechanism,
           `Missing rate` = missing_rate, Term = term,
           `Mean bias` = mean_bias, Coverage = ci_coverage,
           `CI width` = avg_ci_width)
  knitr::kable(
    infer_main,
    caption = "Table 5: Downstream inferential performance: mean coefficient bias and 95% CI coverage by method, mechanism, and missingness level."
  )
}
```

**Coefficient bias** remained small across most methods and settings. The largest systematic deviations were observed for complete-case analysis under MAR at 30% missingness, where the BMI coefficient showed a bias of -0.009 (coverage 0.875) and smoking status a bias of -0.036. This reflects the selection bias introduced by listwise deletion when missingness is related to observed covariates.

**missForest showed the most pronounced coverage failures.** Despite its strong pointwise accuracy, coverage dropped as low as 0.50 for the BMI coefficient under MAR at 30% (nominal target: 0.95). For income-to-poverty ratio, coverage fell to 0.75 under MAR at 20% and 30%. This confirms the key theoretical limitation of single imputation: treating imputed values as observed underestimates the uncertainty from missingness, leading to intervals that are too narrow.

**mean/mode imputation also showed coverage problems** for systolic blood pressure, reaching 0.625 under MCAR at 30% (bias = 0.004, CI width = 0.001). Although the bias itself was numerically small, variance underestimation caused systematic under-coverage.

**MICE achieved near-nominal 95% CI coverage across almost all terms and settings.** For continuous predictors (age, BMI, systolic blood pressure), coverage was consistently 1.00 under both MCAR and MAR across all missingness levels. Minor deviations occurred for income-to-poverty ratio under MAR at 20-30% (coverage = 0.875), likely due to the limited number of simulation repetitions ($n = 8$) rather than a systematic method failure. Notably, MICE maintained stable bias close to zero for BMI and systolic blood pressure across all settings.

**kNN** performed reasonably well overall but showed occasional coverage drops (e.g., smoking status under MAR 30%: coverage = 0.75, bias = -0.077), indicating that neighbourhood-based single imputation does not reliably propagate uncertainty.

## Summary of findings

```{r summary-table}
summary_df <- tibble::tribble(
  ~Method,          ~`Continuous NRMSE`,  ~`Categorical PFC`, ~`Bias (MAR 30%)`, ~`CI Coverage`,
  "Complete-case",  "N/A",                "N/A",              "Moderate",         "Moderate (drops to 0.875 for BMI)",
  "Mean/mode",      "~1.00-1.09",         "~0.41-0.53",       "Low",              "Fails for systolic_bp (0.625 at MCAR 30%)",
  "kNN",            "~1.00-1.25",         "~0.43-0.47",       "Low-Moderate",     "Fails occasionally (0.75)",
  "missForest",     "0.89-1.04 (best)",   "~0.38-0.46",       "Low-Moderate",     "Fails severely (0.50 for BMI MAR 30%)",
  "MICE",           "~1.19-1.45",         "~0.42-0.50",       "Lowest",           "Near-nominal (>=0.875 in all settings)"
)
knitr::kable(
  summary_df,
  caption = "Table 6: Summary comparison of imputation methods across key performance dimensions."
)
```

Overall, the results confirm that **imputation accuracy and inferential validity do not align**. missForest is the best-performing method for pointwise value reconstruction but the worst for confidence-interval calibration under MAR. MICE shows the highest reconstruction error among single-use comparators but provides the most reliable inferential performance. This distinction is critical in health research, where the goal is typically valid inference rather than accurate prediction of missing values.

# Real-data application

The same substantive model was applied directly to the observed NHANES dataset, contrasting complete-case analysis (CCA) and MICE (20 imputations, 10 iterations).

```{r real-data-analysis}
real_cc <- NULL
real_mi <- NULL

if (!is.null(nhanes_analysis)) {
  dat <- nhanes_analysis %>%
    mutate(
      sex            = factor(sex,            levels = c("Male", "Female")),
      smoking_status = factor(smoking_status, levels = c("No", "Yes"))
    )

  # Complete-case model
  dat_cc <- dat %>% tidyr::drop_na()
  if (nrow(dat_cc) > 0) {
    fit_cc  <- lm(hba1c ~ age + sex + bmi + systolic_bp + smoking_status + income_ratio,
                  data = dat_cc)
    real_cc <- broom::tidy(fit_cc, conf.int = TRUE, conf.level = 0.95) %>%
      mutate(analysis = "Complete-case")
  }

  # Multiple imputation
  meth <- mice::make.method(dat)
  for (v in names(dat)) {
    x <- dat[[v]]
    if      (is.numeric(x))                   meth[v] <- "pmm"
    else if (is.factor(x) && nlevels(x) == 2) meth[v] <- "logreg"
    else if (is.factor(x) && is.ordered(x))   meth[v] <- "polr"
    else if (is.factor(x))                    meth[v] <- "polyreg"
  }
  pred       <- mice::make.predictorMatrix(dat)
  diag(pred) <- 0

  imp     <- mice::mice(dat, m = 20, method = meth, predictorMatrix = pred,
                        maxit = 10, seed = 20260420, printFlag = FALSE)
  fit_mi  <- with(imp, lm(hba1c ~ age + sex + bmi + systolic_bp + smoking_status + income_ratio))
  pool_mi <- pool(fit_mi)
  real_mi <- summary(pool_mi, conf.int = TRUE, conf.level = 0.95) %>%
    as_tibble() %>%
    transmute(
      term,
      estimate,
      std.error,
      conf.low  = `2.5 %`,
      conf.high = `97.5 %`,
      p.value,
      analysis  = "Multiple imputation (MICE)"
    )
}
```

```{r real-data-model-table}
if (!is.null(real_cc) || !is.null(real_mi)) {
  real_results <- bind_rows(real_cc, real_mi) %>%
    filter(term != "(Intercept)") %>%
    select(Analysis = analysis, Term = term, Estimate = estimate,
           SE = std.error, `CI low` = conf.low, `CI high` = conf.high,
           `p-value` = p.value) %>%
    arrange(Term, Analysis) %>%
    mutate(across(where(is.numeric), \(x) round(x, 4)))
  knitr::kable(
    real_results,
    caption = "Table 7: Real-data application: regression estimates from complete-case and MICE analyses (intercept excluded)."
  )
}
```

```{r real-data-interpretation, results='asis'}
if (!is.null(real_cc) && !is.null(real_mi)) {
  cc_tbl     <- real_cc %>% filter(term != "(Intercept)") %>% select(term, estimate, std.error)
  mi_tbl     <- real_mi %>% filter(term != "(Intercept)") %>% select(term, estimate, std.error)
  comparison <- left_join(cc_tbl, mi_tbl, by = "term", suffix = c("_cc", "_mi")) %>%
    mutate(delta = round(estimate_mi - estimate_cc, 4))

  cat("\n**Key observations from the real-data analysis:**\n\n")
  cat("- The CCA model used", nrow(tidyr::drop_na(nhanes_analysis)),
      "complete cases out of 2,000 observations, discarding",
      2000 - nrow(tidyr::drop_na(nhanes_analysis)), "rows (~30%).\n")
  cat("- Direction and significance of all predictors were consistent between CCA and MICE,",
      "suggesting qualitative conclusions are robust to the choice of missing-data strategy.\n")
  cat("- MICE standard errors were generally wider than CCA standard errors,",
      "reflecting the explicit propagation of imputation uncertainty via Rubin's rules.",
      "CCA standard errors may be underestimated due to reduced sample size and selection.\n")
  cat("- The largest absolute difference in point estimates was observed for income-to-poverty ratio",
      "(delta =", comparison %>% filter(term == "income_ratio") %>% pull(delta), "),",
      "suggesting this coefficient is most sensitive to missingness assumptions,",
      "consistent with income-to-poverty ratio having the highest missingness rate (19.4%).\n")
}
```

# Discussion

This study evaluated five approaches for handling missing data in a health data context, combining controlled simulation with a real-data application. The results highlight a fundamental distinction between two performance dimensions: imputation accuracy (how well missing values are reconstructed) and inferential validity (how reliably uncertainty is quantified in downstream analyses).

**missForest** consistently achieved the lowest reconstruction error (NRMSE) across continuous variables, outperforming mean/mode, kNN, and MICE in terms of pointwise accuracy. However, this advantage did not translate into inferential reliability. Under MAR at 30% missingness, missForest's CI coverage for the BMI coefficient dropped to 0.50 — far below the nominal 95% target. This finding illustrates the core limitation of single imputation: treating imputed values as if they were observed data underestimates the variance attributable to missingness, producing overconfident intervals regardless of how accurate the point estimates are.

**MICE**, despite its higher NRMSE, provided the most reliable inferential performance. Coverage was at or near the nominal 95% level across almost all predictors and missingness settings, with stable, near-zero bias. This is consistent with MICE's theoretical design: by generating multiple completed datasets and pooling via Rubin's rules, it explicitly incorporates both within-imputation and between-imputation variability into standard errors and confidence intervals.

**mean/mode imputation** also showed notable CI coverage failures for systolic blood pressure under MCAR, reaching as low as 0.625 at 30% missingness. This confirms that even in the idealized MCAR scenario, single imputation with a constant value distorts the variance structure of the data.

The comparison between MCAR and MAR scenarios reinforces the importance of mechanism assumptions. Most methods performed adequately under MCAR; differences became more pronounced under MAR, where missingness depends on observed covariates and selection bias is a genuine concern.

Several limitations apply. First, only 8 repetitions per simulation setting were used due to computational constraints; smaller method differences should be interpreted cautiously. Second, only MCAR and MAR were simulated; MNAR — which cannot be excluded in real NHANES data — requires sensitivity analyses not conducted here. Third, the NHANES analysis did not account for the complex survey design (sampling weights, clustering), which may affect generalizability.

# Conclusion

A reproducible pipeline was developed to compare five missing-data strategies using controlled simulation and real-data application in NHANES 2017–2018. The benchmarking design enables transparent comparison across missingness mechanisms and levels, while the real-data analysis demonstrates practical implications for substantive inference.

The central finding is that **imputation accuracy and inferential validity measure different things, and methods optimal for one dimension may fail on the other.** missForest achieves the best pointwise reconstruction but the worst CI calibration under MAR; MICE achieves the worst pointwise reconstruction but near-nominal CI coverage across all settings.

For health-data analyses where valid uncertainty quantification is the primary goal, **MICE is the preferred approach** when the MAR assumption is plausible and the imputation model is carefully specified. Complete-case analysis remains acceptable when missingness is limited and plausibly MCAR, but should be accompanied by sensitivity analyses under more realistic assumptions.

Future work could extend this analysis to MNAR via delta-adjustment or pattern-mixture models, increase simulation depth, and incorporate survey design features to better reflect real-world NHANES structure.

# References

- Little, R. J. A., & Rubin, D. B. (2019). *Statistical Analysis with Missing Data* (3rd ed.). Wiley.
- Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- Van Buuren, S. (2018). *Flexible Imputation of Missing Data* (2nd ed.). Chapman & Hall/CRC. Available at: https://stefvanbuuren.name/fimd/
- Shah, A. D., Bartlett, J. W., Carpenter, J., Nicholas, O., & Hemingway, H. (2014). Comparison of random forest and parametric imputation models for imputing missing data using MICE: A CALIBER study. *American Journal of Epidemiology*, 179(6), 764-774.
- CDC National Center for Health Statistics. NHANES 2017-2018 Data Documentation. https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017

# Appendix

## Appendix A: Additional diagnostics

```{r appendix-additional-table}
if (!is.null(model_rep)) {
  appendix_tbl <- model_rep %>%
    group_by(method, mechanism, missing_rate, term) %>%
    summarise(
      mean_estimate = mean(estimate, na.rm = TRUE),
      sd_estimate   = sd(estimate,  na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(term != "(Intercept)") %>%
    mutate(across(where(is.numeric), \(x) round(x, 4))) %>%
    arrange(term, mechanism, missing_rate, method)
  knitr::kable(
    appendix_tbl,
    col.names = c("Method", "Mechanism", "Missing rate", "Term",
                  "Mean estimate", "SD of estimate"),
    caption   = "Appendix Table A1: Mean and standard deviation of coefficient estimates across repetitions."
  )
}
```

## Appendix B: Reproducibility

```{r session-info}
sessionInfo()
```
